Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2013-05-11 paper.tex
typo in equation
more updates

The presence of $q_i$ in this expression serves as a reference state.
The idea is that the error in the characteristic projection of the
jump (due to the nonlinearity) is minimized if we pick a suitable
reference state, since we only care about the jumps carried to the
interface over the timestep \citep{colellaglaz1985}.  The original
Castro implementation used the cell-average quantity.  Experiments
found that this is ill-behaved in the presence of strong shocks
(especially at low resolution). For the present work, we switch the
reference state to
\begin{equation}
\tilde{q}_L = \left \{ \begin{array}{cc}
       \mathcal{I}_+^{(+)}(q_i) & \mathrm{if~} u + c > 0 \\
       q_i                    & \mathrm{otherwise}
\end{array}
\right .
\end{equation}
where the $(+)$ superscript here means the fastest wave moving to the right
(the $u+c$ eigenvalue).   This is simply the average under the largest
portion of the parabolic profile that could possible reach the interface 
over the timestep.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  This makes our
expression appear as:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q_i}  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
The final change is that now we can no longer simply multiply
$\tilde{q_i} - \mathcal{I}^{(\nu)}_+(q_i)$ by the flattening
parameter, $\chi$, but instead blend the traced state with the cell
centered state as:
  \begin{equation}
  q_{i+1/2,\{L,R\}}^{n+1/2} \leftarrow (1 - \chi) q_i + \chi q_{i+1/2,\{L,R\}}^{n+1/2}
  \end{equation}
This is the same prescription used in the \cite{ppm}.

This change can be used in the public version of Castro by setting
{\tt castro.ppm\_reference = 1}.  Figure~\ref{Fig:sod} shows the
solution to the Sod problem using the old and new reference states.
We see that the shock and contact are slightly sharper with this new
reference state, but also there is a slight dip at the tail of the
rarefaction with the new method.

@@

We note a few additional differences between the original PPM
implementation of \citet{ppm} and Castro.  In the original
implementation, the gravitational source term was included in the
characteristic projection.  Here, we follow \citet{ppmunsplit} and add
it afterwards.  In our experiments, projecting the source term did not
make a noticable difference in the energy conservation of the
solution.  Finally, the PPM paper uses an edge-centered gravitational
acceleration in the prediction, whereas we use a cell-centered one.
Both are second-order accurate.  In the unsplit implementation, there
is an additional freedom that comes into play---do we add the source
terms to the normal predictor before or after using it to construct
the transverse flux differences?  Our experiments showed no
substantial difference with either variant.  We add the sources only
at the final construction of the interface states.

The boundary conditions on the hyperbolic system are simply
zero-gradient.  Using AMR, we make the coarse grid very large, so as
to place the boundaries far from the region of interest.  For the
gravity solve, we would like isolated boundary conditions.  Here we
experiment with using two monopoles (one for each star).  The
individual potentials are summed and this is evaluated on the physical
boundary to set Dirichlet conditions for the Poisson solve.

@@

>>>>

The presence of $q_i$ in this expression serves as a reference state.
The idea is that the error in the characteristic projection of the
jump (due to the nonlinearity) is minimized if we pick a suitable
reference state, since we only care about the jumps carried to the
interface over the timestep \citep{colellaglaz1985}.  The original
Castro implementation used the cell-average quantity.  Experiments
found that this is ill-behaved in the presence of strong shocks
(especially at low resolution). For the present work, we switch the
reference state to
\begin{equation}
\tilde{q}_L = \left \{ \begin{array}{cc}
       \mathcal{I}_+^{(+)}(q_i) & \mathrm{if~} u + c > 0 \\
       q_i                    & \mathrm{otherwise}
\end{array}
\right .
\end{equation}
where the $(+)$ superscript here means the fastest wave moving to the right
(the $u+c$ eigenvalue).   This is simply the average under the largest
portion of the parabolic profile that could possible reach the interface 
over the timestep.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  This makes our
expression appear as:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
The final change is that now we can no longer simply multiply
$\tilde{q_i} - \mathcal{I}^{(\nu)}_+(q_i)$ by the flattening
parameter, $\chi$, but instead blend the traced state with the cell
centered state as:
  \begin{equation}
  q_{i+1/2,\{L,R\}}^{n+1/2} \leftarrow (1 - \chi) q_i + \chi q_{i+1/2,\{L,R\}}^{n+1/2}
  \end{equation}
This is the same prescription used in the \cite{ppm}.

This change is now the default in the public version of Castro.
Figure~\ref{Fig:sod} shows the
solution to the Sod problem using the old and new reference states.
We see that the shock and contact are slightly sharper with this new
reference state, but also there is a slight dip at the tail of the
rarefaction with the new method.

@@

We note a few additional differences between the original PPM
implementation of \citet{ppm} and Castro.  In the original
implementation, the gravitational was reconstructed as a parabola,
and this was traced under to find the forcing that affects the interface
for each wave.   Here, we follow \citet{ppmunsplit} and add   \MarginPar{we have a version that does the tracing---are we using it?  But there is a $\Delta t/2$ difference still}
it afterwards.  In our experiments, projecting the source term did not
make a noticable difference in the energy conservation of the
solution.  In the unsplit implementation, there
is an additional freedom that comes into play---do we add the source
terms to the normal predictor before or after using it to construct
the transverse flux differences?  Our experiments showed no
substantial difference with either variant.  We add the sources only
at the final construction of the interface states.

The boundary conditions on the hyperbolic system are simply
zero-gradient.  Using AMR, we make the coarse grid very large, so as
to place the boundaries far from the region of interest.  We further
make the restriction that only level-0 (coarse) grids can touch the
domain boundary.

@@

For the
gravity solve, we would like isolated boundary conditions.  Here we
experiment with using two monopoles (one for each star).  The
individual potentials are summed and this is evaluated on the physical
boundary to set Dirichlet conditions for the Poisson solve.
